### 2017-01-12 Manuscript/SW_early_life_telomeres.Rmd
Home edits

2. Here we study spatio-temporal variation in lifelong telomere dynamics in the Seychelles warbler (*Acrocephalus sechellensis*). We combine long-term ecological data with one the largest longitudinal telomere datasets to date, consisting of `r nrow(dd)` samples from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`. We provide a detailed analysis of how telomere dynamics vary over individual lifespans, and with spatio-temporal variation in the social and ecological environment.

3. We found that telomere length decreases with cross-sectional and longitudinal measures of age, and that telomere length decreases most rapidly early in life. However, both cross-sectional and longitudinal data suggested that against this overall pattern of shortening, bouts of telomere length increases occur in some individuals. Using a large number of repeated measurements we show statistically that these increases cannot be explained solely by qPCR measurement error.

4. Telomere length decreased with age in almost all cohorts studied, but telomere length and the slope of the relation ship with age varied markedly among cohorts. Variation in telomere length was positively associated with island-wide temporal variation in insect abundance, suggesting that temporal variation in food availability imposes a cost in Seychelles warblers. We also found evidence for sex-specific relationships between telomeres and tarsus length, potentially reflecting differential costs of growth.

5. Our long-term data show that in natural populations, telomere dynamics vary in a highly complex manner over individual lifespans, and across space and time. Variance in telomere dynamics among individuals is the product of wide array of genetic, parental and environmental factors. Explaining this variation more fully will require the integration of comprehensive long-term ecological and genetic data from multiple populations and species.

**Keywords:** Life-history; Telomere; Seychelles warbler; Senescence

@@

###Introduction
A major aim of ecologists is to be able to quantify and understand why individuals vary in their response to different environments. Identifying the costs imposed on individuals within populations is central to understanding variation in fitness [@Lindstrom1999], and thus for understanding population and community dynamics [@Bolnick2011]. However, fully quantifying individual-level variation in costs is impossible in wild systems, and thus effective biomarkers that reflect the physiological consequences of individual-level experiences are required.

Telomeres are repetitive DNA sequences found on the ends of linear chromosomes that protect against DNA damage. Telomeres shorten with age [@Monaghan2006], and there is evidence from a range of taxa that the greatest rate of telomere shortening occurs in early life [e.g. @Frenck1998; @Haussmann2003]. Telomere shortening also occurs in response to oxidative stress, which can be elevated due to environmental factors [@VonZglinicki2002]. There is evidence from both humans and wild animal populations that telomere shortening is influenced by the conditions experienced during both early life and adulthood [@Price2013; @Monaghan2014; @Nettle2015; @Reichert2015]. Importantly, the extent of telomere shortening that occurs has a direct link to senescence and survival. When telomeres become critically short cells senesce [@Campisi2003], and the accumulation of these cells can result in organismal senescence and death [@Wong2003]. This association between senescence and telomere length has inspired a great deal of recent research into telomere evolutionary ecology, and relationships between telomere dynamics and survival have been documented in several wild organisms [reviewed in @Horn2010; @Haussmann2010; @Simons2015]. While there is as yet little direct evidence that the relationship between telomere dynamics and survival is causal [@Simons2015], there is mounting evidence that telomeres can act as biomarkers of cost in wild populations, providing a signature of the ecological stress that has been experienced and is otherwise difficult to detect [e.g. @Schultner2014; @Asghar2015].

Although there has now been a considerable amount of effort put into studying telomere dynamics in natural populations, our understanding of the forces responsible for explaining variation in telomere length is still limited. This is important, as before we can use telomeres as a measure of the costs experienced by individuals and populations, we need to know how different developmental, genetic and ecological variables interact to affect telomeres. Telomere length and rates of shortening can vary according to parental characteristics [Njajour2007; @Heidinger2016], among sexes [@Barrett2011], and with a whole host of environmental conditions experienced at different life-history stages [@Monaghan2014]. There is also some evidence that telomeres can increase, as well as decrease, in length [@Simons2014], which has important ramifications for our understanding of how telomeres reflect costs. However, increases in telomere length are often attributed to measurement error [@Steenstrup2013], and as such its ecological significance is unknown. Recent evidence suggests that telomere dynamics are indeed highly variable over individual lifespans, and that even the relationship between telomeres and age can vary markedly among cohorts [@Fairlie2016]. To better understand what explains variation in telomere dynamics, more studies are required that incorporate telomere variation over entire lifespans with comprehensive, long-term ecological data.

The longitudinal study (since 1986) of the Seychelles warbler (*Acrocephalus sechellensis*) population on Cousin Island provides an excellent system for studying senescence in the wild [reviewed in @Hammers2015]. Due to the isolated nature of the study population and intensive field monitoring, we have unusually comprehensive ecological and survival data spanning many years (see Methods, below). Conditions and warbler population density on Cousin vary across space and time due to weather-induced changes in foliage cover and food availability [@VandeCrommenacker2011a]. Variation in the oxidative stress experienced by individuals is associated with territory quality [@VandeCrommenacker2011a]. However, neither early-life nor adult survival appear to be associated with territory quality or local density [@Brouwer2006; @Hammers2013]. Facultative cooperative breeding occurs in the Seychelles warbler [@Komdeur1994a; @Richardson2003a], and the presence of helping subordinates (but not non-helping subordinates) in the natal territory is associated with increased survival later in life [@Brouwer2012]. Lastly, we have an established protocol for assessing telomere length in this species [@Barrett2012; @Bebbington2016], and telomere length predicts survival independently of age, suggesting that telomeres act as a biomarker of cost in this species [@Barrett2013a; @Hammers2015; @Bebbington2016]. Thus, we have an excellent system in which to assess the costs of different social and environmental conditions experienced by individuals, and to assess how these costs vary over space and time.

In this study we study how lifelong telomere dynamics are related to environmental variation across `r length(table(dd$Cohort))` Seychelles warbler cohorts. We first study how telomere length and rates of shortening are related to age, and how this relationship varies among cohorts, in order to gain an in-depth understanding of the temporal dynamics of telomere changes. We then examine, within individuals, how telomere length changes with age, and statistically test whether increases in telomere length are biologically meaningful. Finally, we test how a wide range of social and environmental variables are related to telomere length and shortening in order gain a fuller understanding of the forces driving telomere dynamics in natural populations.

@@

#####Study species and sampling
The Seychelles warbler is a small (~15 g), insectivorous passerine bird with a mean life expectancy of 5.5 years at fledging [@Hammers2013]. The population of *ca*. 320 birds on Cousin Island (04'20'S, 55'40'E) has been intensively studied since 1986 [@Richardson2003b; @Spurgin2014a]. This species' main breeding season runs from June-August, when the breeding females on each of the *ca*. 115 territories lay one or, rarely, two or three eggs [@Komdeur1991]. As a result of this low reproductive output, combined with higher mortality in first-year birds [39%; @Brouwer2006], cohort sizes in the Seychelles warbler are typically small (< 50).

Individuals are usually ringed in their first year of life, and so are of known age. They are then followed throughout their lives, and as they are non-migratory endemics naturally confined to the island [@Komdeur2004a], a biannual census of birds on Cousin during each breeding season gives accurate measures of local density, social status (e.g. breeding male/female, helping subordinate, non-helping subordinate) and individual survival [@VandeCrommenacker2011; @Barrett2013a]. Full details of monitoring methods can be found in Brouwer *et al.* [-@Brouwer2012].

Seychelles warblers are highly territorial and all territories are mapped during the breeding seasons using detailed observational data of foraging and territorial defence behaviour, and surveyed for territory quality [@Richardson2003b]. Territory quality is calculated based on territory size, foliage cover and insect abundance [@Komdeur1992], and where territory quality estimates were not available for a specific year we used the average value for that territory across years [@Hammers2013]. Cousin is subject to intra- and inter-annual variation in rainfall and food availability, and such island-wide temporal variation may override the effects of individual territory quality. As an estimate of seasonal variation in food availability, we calculated an index of the number of insects across the entire island during each breeding season (referred to hereafter as 'insect abundance'). This index is calculated as the mean number of insects found per unit leaf area over all surveys carried out on the island in a breeding season.

Each time a bird is caught on Cousin body mass and tarsus length are measured (to the nearest 0.1g and 0.1mm, respectively). We aged all birds using information on eye colour [@Komdeur1991a] and previous captures [@Richardson2003b]. A blood sample (*ca* 25 $\mu$l) is taken from each bird captured via brachial venipuncture, and stored at room temperature in 1 ml of absolute ethanol in a 1.5 ml screw-cap microfuge tube.

>>>>

2. Here we study spatio-temporal variation in lifelong telomere dynamics in the Seychelles warbler (*Acrocephalus sechellensis*). We combine long-term life history and ecological data with one the largest longitudinal telomere datasets to date, consisting of `r nrow(dd)` samples from `r length(table(dd$Cohort))` cohorts born between `r min(dd$LayYear)` and `r max(dd$LayYear)`. We provide a detailed analysis of how telomere dynamics vary over individual lifespans and cohorts, and with spatio-temporal variation in the social and ecological environment.

3. We found that telomere length decreases with cross-sectional and longitudinal measures of age, and that telomere length decreases most rapidly early in life. However, both cross-sectional and longitudinal data suggested that against this overall pattern of shortening, bouts of telomere length increases occur in some individuals. Using a large number of repeated measurements we show statistically that these increases are unlikely to be explained solely by qPCR measurement error.

4. Telomere length decreased with age in almost all cohorts studied, but telomere length and the slope of the relation ship with age varied markedly among cohorts. Telomere length was positively associated with island-wide temporal insect abundance, suggesting that the costs associated with living in a harsher environment can be studied by investigating telomere dynamics. We also found evidence for sex-specific relationships between telomeres and tarsus length, potentially reflecting differential costs of growth.

5. Our long-term data show that in a natural population, telomere dynamics vary in a highly complex manner over individual lifespans, and across space and time. Variance in telomere dynamics among individuals is the product of wide array of genetic, parental and environmental factors. Explaining this variation more fully will require the integration of comprehensive long-term ecological and genetic data from multiple populations and species.

**Keywords:** Biomarkers; Intra- and inter-individual variation; Life-history; Telomere; Seychelles warbler; Senescence

@@

###Introduction
A major aim of ecologists is to quantify and understand why individuals vary in their response to different environments. Identifying the costs imposed on populations is central to understanding variation in fitness [@Lindstrom1999], and thus for understanding population and community dynamics [@Bolnick2011]. However, fully quantifying individual-level variation in costs is impossible in wild systems, and thus effective biomarkers that reflect the physiological consequences of individual-level experiences are required.

Telomeres have been proposed to be a potential biomarker of such costs [@Monaghan2014]. Telomeres are repetitive DNA sequences on the ends of linear chromosomes that protect against DNA damage. Telomeres generally shorten with age [@Monaghan2006], and there is evidence from a range of taxa that the greatest rate of telomere shortening occurs in early life [e.g. @Frenck1998; @Haussmann2003]. Telomere shortening occurs in response to oxidative stress, which can be elevated due to environmental factors [@VonZglinicki2002]. There is evidence from both humans and wild animal populations that telomere shortening is influenced by the conditions experienced during both early life and adulthood [@Price2013; @Monaghan2014; @Nettle2015; @Reichert2015]. Importantly, the extent of telomere shortening is directly linked to senescence and survival. When telomeres become critically short cells senesce [@Campisi2003], and the accumulation of these cells can result in organismal senescence and death [@Wong2003].

The association between senescence and telomere length has inspired a great deal of recent research into telomere evolutionary ecology, and relationships between telomere dynamics and survival have been documented in several wild organisms [reviewed in @Horn2010; @Haussmann2010; @Simons2015]. While there is as yet little direct evidence that the relationship between telomere dynamics and survival is causal [@Simons2015], there is mounting evidence that telomeres can act as biomarkers of cost in wild populations, providing a signature of the ecological stress that has been experienced and is otherwise difficult to detect [e.g. @Schultner2014; @Asghar2015]. There is also some evidence that telomeres can increase, as well as decrease, in length [@Simons2014; @Bateson2016], which has important ramifications for our understanding of how telomeres reflect costs. However, increases in telomere length are often attributed to measurement error [@Steenstrup2013], and as such its ecological significance is unknown.

Although a considerable amount of effort has been put into studying telomere dynamics in natural populations, our understanding of the forces responsible for explaining variation in telomere length is still limited. Understanding how different factors shape telomere length variation is important, as before we can use telomeres as a measure of the costs experienced by individuals, we need to know how different developmental, genetic and ecological variables interact to affect telomeres. Telomere length and rates of shortening can vary according to parental characteristics [Njajour2007; @Heidinger2016], among sexes [@Barrett2011], and with a whole host of environmental conditions experienced at different life-history stages [@Monaghan2014]. Recent evidence suggests that telomere dynamics are indeed highly variable over individual lifespans, and that even the relationship between telomeres and age can vary markedly among cohorts [@Fairlie2016]. To understand which factors best explain variation in telomere dynamics, more studies are required that incorporate telomere variation over entire lifespans with comprehensive, long-term ecological data.

The longitudinal study (since 1986) of the Seychelles warbler (*Acrocephalus sechellensis*) population on Cousin Island provides an excellent system for studying telomere dynamics and senescence patterns in the wild [reviewed in @Hammers2015]. Due to the isolated nature of the study population and intensive field monitoring, we have comprehensive ecological and survival data spanning many years (see Methods, below). Environmental conditions and population density on Cousin Island vary across space and time due to weather-induced changes in foliage cover and food availability [@VandeCrommenacker2011a]. Variation in oxidative stress experienced by individuals is associated with territory quality [@VandeCrommenacker2011a]. However, neither early-life nor adult survival appear to be associated with territory quality or local density [@Brouwer2006; @Hammers2013].

 As well as territory quality variation, there is also variation in the social environment. Facultative cooperative breeding occurs in the Seychelles warbler [@Komdeur1994a; @Richardson2003a], and the presence of helpers (but not non-helpers) in the natal territory is associated with increased survival of offspring later in life [@Brouwer2012]. Lastly, we have an established protocol for assessing telomere length in this species [@Barrett2012; @Bebbington2016], and telomere length predicts survival independently of age, suggesting that individual variation in telomere length is ecologically relevant in this species [@Barrett2013a; @Hammers2015; @Bebbington2016]. Thus, we have an excellent system in which to assess the costs of different social and environmental conditions experienced by individuals, and to assess how these costs vary over space and time.

In this study, we test how lifelong telomere dynamics are related to environmental variation across `r length(table(dd$Cohort))` Seychelles warbler cohorts. We first study how telomere length and rates of shortening are related to age, and how this relationship varies among cohorts, in order to gain an in-depth understanding of the temporal dynamics of telomere changes. We then examine, within individuals, how telomere length changes with age, and statistically test whether observed increases in telomere length are larger than expected based on measurement error. Finally, we test how a wide range of social and environmental variables are related to telomere length and shortening in order gain a fuller understanding of the forces driving telomere dynamics in natural populations.

@@

#####Study species and sampling
The Seychelles warbler is a small (~15 g), insectivorous passerine bird with a mean life expectancy of 5.5 years at fledging [@Hammers2013]. The population of *ca*. 320 birds on Cousin Island (04'20'S, 55'40'E) has been intensively studied since 1986 [@Richardson2003b; @Spurgin2014a]. This species' main breeding season runs from June-September, when the breeding females on each of the *ca*. 115 territories lay one or, rarely, two or three eggs [@Komdeur1991]. As a result of this low reproductive output, combined with higher mortality in first-year birds [39% in first-year birds versus 16% in adults; @Brouwer2006], cohort sizes in the Seychelles warbler are typically small (< 50).

Individuals are usually ringed (with an individually numbered metal ring and unique combination of colour rings) in their first year of life, and so are of known age. We aged all birds using information on eye colour [@Komdeur1991a] and previous captures [@Richardson2003b]. As Seychelles warblers are non-migratory endemics naturally confined to the island [@Komdeur2004a], a biannual census of birds on Cousin during each breeding season gives accurate measures of local density, social status (e.g. breedier, helper, non-helper) and individual survival [@VandeCrommenacker2011; @Barrett2013a]. Full details of monitoring methods can be found in Brouwer *et al.* [-@Brouwer2012].

Seychelles warblers are highly territorial and all territories were mapped during each main breeding season using detailed observational data of foraging and territorial defence behaviour, and surveyed for territory quality [@Richardson2003b]. Territory quality is calculated based on territory size, foliage cover and insect abundance [@Komdeur1992]. Where territory quality estimates were not available for a specific year we used the average value for that territory across years [@Hammers2013]. Cousin is subject to intra- and inter-annual variation in rainfall and food availability, and such island-wide temporal variation may override the effects of individual territory quality. As an estimate of seasonal variation in food availability, we calculated an index of the number of insects across the entire island during each main breeding season (referred to hereafter as 'insect abundance'). This index is calculated as the mean number of insects found per unit leaf area over all surveys carried out on the island in a main breeding season.

Each time a bird is caught on Cousin body mass and tarsus length are measured (to the nearest 0.1g and 0.1mm, respectively). A blood sample (*ca* 25 $\mu$l) is taken from each bird captured via brachial venipuncture, and stored at room temperature in 1 ml of absolute ethanol in a 1.5 ml screw-cap microfuge tube.
